Pillar 1 - Sequential Pattern Mining

### 2019-01-30 Pillar 1 - Sequential Pattern Mining.Rmd
Text editing

# Setup
load previous data and create function to find courses
```{r loading data}
load("Output/data_pillar_1.RDATA")
```

Function that returns title of a course given its code.
```{r function find_course}
find_course <- function(code){

@@

# Data Exploration

```{r statistic summary}

# For convenience
provide_statistics <- function(data){
  data %>%
    summarise(
      Min       = min(Grade),
      Max       = max(Grade), 
      Mean      = round(mean(Grade), 2), 
      Median    = median(Grade), 
      SD        = round(sd(Grade), 2),
      Fail_rate = round(mean(Fail), 2),
      Fail_n    = sum(Fail),
      n         = n()
      )
}

@@

#
# output
save(statistics_student, statistics_course, statistics_cluster,
     statistics_concentration, statistics_year, statistics_level, file = "Output/Transcript Statistics.RDATA")
rm(provide_statistics, statistics_student, statistics_course, statistics_cluster,
     statistics_concentration, statistics_year, statistics_level)
```

@@

### CSPADE Algorithm
```{r}
# Transform rules from ruleInduction into a readable data frame
n_students <- length(unique(d_transactions$sequenceID))

clean_SR<- function(rules){

  rules %>%
    as("data.frame") %>%
    mutate(
      rule = str_remove_all(rule, pattern = "[<]"),
      rule = str_remove_all(rule, pattern = "[{]"),
      rule = str_remove_all(rule, pattern = "[}]"),
      rule = str_remove_all(rule, pattern = "[>]")
      ) %>%
    separate(rule, into = c("lhs", "rhs"), sep="=")  %>%
    select(
      lhs, rhs,
      support,
      confidence,
      lift
    ) %>%
    separate(lhs, into=c("LHS Course ID", "LHS Quality"), sep="_", remove= F) %>%
    separate(rhs, into=c("RHS Course ID", "RHS Quality"), sep="_", remove= F) %>%
    # Compute statistic
    group_by(lhs,`RHS Course ID`) %>%
    mutate(
      lhs.rhsCourse.support = sum(support),
      confidence.corr = support / lhs.rhsCourse.support,
      count = support * n_students,
      lhs.rhsCourse.count = lhs.rhsCourse.support * n_students

      # TODO: coñpute rhs.support
      # lift.corr = confidence.corr / rhs.support

  ) %>%
  ungroup() %>%
    mutate_at(
      c("support",
        "confidence", "lift"),
      funs(round(., 5))
    )

```
We apply cspade followed by rule induction
```{r}
# Run the cspade algorithm with desired parameters. Then run the ruleInduction algorithm with desired parameters.
my_SR<- function(data){
 data %>%
  cspade(
    parameter = list(
      support = 0, # min support of a sequence
      maxsize = 1, # max number of items of an element of a sequence
      maxlen  = 2, # max number of element of a sequence
      mingap  = 1, # min time difference between consecutive element of a sequence
      maxgap  = 1e4 # max time difference between consecutive element of a sequence
      #maxwin  = 1e4  # max time difference between any two elements of a sequence # WARNING: 'maxwin' disabled
      ),
    control = list(
      verbose = FALSE
      )
    ) %>%
    ruleInduction(
      confidence = 0, #we need to play with the parameters.
      control    = list(verbose = FALSE)
      ) %>%
    clean_SR
}

#Generating sequence rules
SR   <- my_SR(data = sequences   )
SR_PF <- my_SR(data = sequences_PF)
SR_HL <- my_SR(data = sequences_HL)

#Saving sequence rules
SR <- list(
    SR = SR,
    SR_PF = SR_PF,
    SR_HL = SR_HL
    )

save(SR, file = "App/SR.RDATA")

#clean unnecessary objects
rm(clean_SR, my_SR, 
   sequences, sequences_PF, sequences_HL, 
   n_students,
   SR, SR_PF, SR_HL)

```
#Experiments
#Visualizing rules
```{r}
#look at counts
ggplot(for_plots, aes(LHS))+
  geom_bar()

@@

##TO MOVE UP (These are ready:)
Rounds grades and add new row from original grade up to 10.
NOTE: includes mandatory courses.
```{r}

distributed_transcript <- d_transcript %>%
  filter(!is.na(Grade), Grade>=0) %>%
  mutate(Round_grade = paste("g", round(Grade,0),sep=""),
         Values = T) %>%
  spread(key = "Round_grade", value = Values) %>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new) #%>%
  #to check that the grades appear from original grade to 10.
  #arrange(`Student ID`, `Course ID`)

>>>>

# Setup
We load the environment `data_pillar1` which we saved at the end of the data preparation. It contains the data sets `d_course` and `d_transcript`.
```{r loading data}
load("Output/data_pillar_1.RDATA")
```

We create a function, which, when given the code of a course, returns its title.
```{r function find_course}
find_course <- function(code){

@@

# Data Exploration
We compute summary statistics (minimum, maximum, mean, median, standard deviation, failure rate, number of failure and count) at different levels (student, course, cluster, concentration, year and course level). We save the results in the environment `Transcript Statistics`.
```{r statistic summary}

# For convenience
provide_statistics <- function(data){

  data %>%
    summarise(
      Min       = min(Grade),
      Max       = max(Grade), 
      Mean      = mean(Grade), 
      Median    = median(Grade), 
      SD        = sd(Grade),
      `Failure Rate` = mean(Fail),
      `Failure Count`    = sum(Fail),
      Count     = n()
      ) %>%
    mutate_at(
      vars(Mean, SD, `Failure Rate`),
      round,
      digits = 2
    )

@@

#
# output
save(statistics_student, statistics_course, statistics_cluster,
     statistics_concentration, statistics_year, statistics_level, file = "Output/Transcript Statistics.RDATA")

rm(provide_statistics, statistics_student, statistics_course, statistics_cluster,
     statistics_concentration, statistics_year, statistics_level)
```

@@

### CSPADE Algorithm
We apply cspade algorithm
```{r}
# Transform rules from ruleInduction into a readable data frame
clean_IR <- function(rules){
  rules %>%
    as("data.frame") %>%
    # filter(
    #   count >= 1
    # ) %>%
    mutate(
      rule = str_remove_all(rule, pattern = "[<]"),
      rule = str_remove_all(rule, pattern = "[{]"),
      rule = str_remove_all(rule, pattern = "[}]"),
      rule = str_remove_all(rule, pattern = "[>]"),
      `rhs.support` = confidence / lift
      ) %>%
    separate(rule, into = c("lhs", "rhs"), sep="=") %>%
    mutate_at(
      c("support", #"lhs.support",
        "confidence", "rhs.support", "lift"),
      funs(round(., 5))
    ) %>%
    select(
      lhs, rhs,
      support, 
      #count, lhs.support, #figure out how to get these
      confidence, rhs.support,
      lift
    ) #%>%
    #arrange(
    #  desc(count)
    #)

}
```
We apply cspade followed by rule induction
```{r}
# Run the cspade algorithm with desired parameters. Then run the ruleInduction algorithm with desired parameters.
my_IR <- function(data){
  data %>%
  cspade(
    parameter = list(
      support = 0.01, # min support of a sequence
      maxsize = 1, # max number of items of an element of a sequence
      maxlen  = 2, # max number of element of a sequence
      mingap  = 1, # min time difference between consecutive element of a sequence
      maxgap  = 1e4 # max time difference between consecutive element of a sequence
      #maxwin  = 1e4  # max time difference between any two elements of a sequence # WARNING: 'maxwin' disabled
      ),
    control = list(
      verbose = FALSE
      )
    ) %>%
    ruleInduction(
      confidence = 0, #we need to play with the parameters.
      control    = list(verbose = FALSE)
      ) %>%
    clean_IR
}

#Generating induced rules
IR    <- my_IR(data = sequences   )
IR_PF <- my_IR(data = sequences_PF)
IR_HL <- my_IR(data = sequences_HL)

#Saving induced rules:
save(
  IR = list(
    IR = IR,
    IR_PF = IR_PF,
    IR_HL = IR_HL
    ),
  file = "Output/IR.RDATA")

rm(clean_IR, my_IR,
   sequences, sequences_PF, sequences_HL,
   IR, IR_PF, IR_HL)

##calculate appropriate measures:
```{r}
n_students <- length(unique(d_sequence_transactions$`Student ID`))

sequence_rules <- pass_rules_df %>%
  separate(LHS, into=c("LHS Course ID", "LHS Score"), sep="_", remove= FALSE) %>%
  separate(RHS, into=c("RHS Course ID", "RHS Score"), sep="_", remove= FALSE) %>%
  mutate(`RHS Course ID`= substr(`RHS Course ID`, start=4, stop=10))%>%
  left_join(lookup_table, by = c("RHS Course ID"="Course ID")) %>%

  # Compute statistics
  group_by(LHS,`RHS Course ID`) %>%
  mutate(Prob_both = sum(support), 
         Conf_corr = support/Prob_both,
         n_rule = support * n_students,
         n_both =Prob_both * n_students,
         lift_diff= Conf_corr - Prop_low,
         lift_corr=Conf_corr / Prop_low) %>%
  ungroup() %>%

  # Filter and edit output
  filter(str_detect(`RHS Score`, pattern = "low"),
         str_detect(`LHS Score`, pattern = "low"),
         n_rule >= 5,
         Conf_corr > 0.2) %>%
  select(LHS, RHS, support, n_rule, Prob_both, n_both, Prop_low, Conf_corr, lift_corr, lift_diff) %>%
  mutate(support = round(support,3),
         Prob_both =round(Prob_both,3),
         Prop_low =round(Prop_low,3),
         Conf_corr= round(Conf_corr,3),
         lift_diff= round(lift_diff,3),
         lift_corr= round(lift_corr,3)) %>%
  arrange(lift_diff)

```
##Save data
```{r}
save(sequence_rules,
     file = "sequence_rules.RDATA")
```

##Arrange data for plots:
```{r}
for_plots <-  manual_confidence_rules %>%
  group_by(LHS)%>%
  mutate(LHS_count=n())
```
#Visualizing rules
```{r}
#look at counts
ggplot(for_plots, aes(LHS))+
  geom_bar()

@@

##TO MOVE UP (These are ready:)
Rounds grades and add new row from original grade up to 10.
NOTE: includes mandatory courses.
```{r}
distributed_transcript <- d_transcript %>%
  filter(!is.na(Grade), Grade>=0) %>%
  mutate(Round_grade = paste("g", round(Grade,0),sep=""),
         Values = T) %>%
  spread(key = "Round_grade", value = Values) %>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10,
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new) #%>%
  #to check that the grades appear from original grade to 10.
  #arrange(`Student ID`, `Course ID`) 
```

##Not ready to move up:
